TITLE by LAURA HOYTE
========================================================

```{r Load and install packages, echo=FALSE, message=FALSE, warning=FALSE, packages}
# Data sourced from the New York City Taxi and Limousine Commission
# Data loading and cleansingg takes place in this file

install.packaage(rmongodb)
install.packages(sqldf)
install.packages(mongolite)

library(rmongodb)
library(sqldf)
library(ggplot2)
library(mongolite)
library(dplyr)
library(jsonlite)
library(lubridate)

taxidata <- read.csv("taxidata3_loc.csv", header = TRUE, stringsAsFactors = T)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data

# Get 10% sample of yellow taxi data
rowcount <- read.csv.sql("yellow_tripdata_2015-05.csv", sql = "select count(*) from file")
yellow <- read.csv.sql("yellow_tripdata_2015-05.csv", sql = "select * from file order by random() limit 1315826")

# Write yellow sample data frame to csv file
write.csv(yellow, file = "yellow_samples.csv", quote = TRUE, na = "NA", row.names = FALSE,
          col.names = TRUE)

# Get 10% sample of green taxi data
#options(warn=3)
rowcount <- read.csv.sql("green_tripdata_2015-05.csv", sql = "select count(*) from file")

# Error when reading in data using the above - more columns than column names. Read in data only, no headers
green <- read.csv("green_tripdata_2015-05.csv", header = FALSE, stringsAsFactors =  FALSE, check.names = F, skip = 1)

# Get all of the data including header
data_and_header <- readLines("green_tripdata_2015-05.csv")

#Filter out header
headers <- strsplit(data_and_header[1], ",")[[1]]

# Add header to csv file
colnames(green) <- headers

# Drop errorneous columns
length(green)
green$"NA" <- NULL
green$"NA" <- NULL

# Check structure and length of green data frame
str(green)
length(green)

# Write green data frame to new csv file
write.csv(green, file = "green_datanew.csv", quote = TRUE, na = "NA", row.names = FALSE)

# Get 10% sample of new green taxi data
rowcount <- read.csv.sql("green_datanew.csv", sql = "select count(*) from file")
rowcount
greennew <- read.csv.sql("green_datanew.csv", sql = "select * from file order by random() limit 178685")

# Rename Trip_data column to remove "." - better for MongoDB
names(greennew)[21]<- "Trip_data"

# Write green sample data frame to csv file
write.csv(greennew, file = "green_samples.csv", quote = TRUE, na = "NA", row.names = FALSE)

# Write to new MongoDB collection yellowsample in taxi database
if (mongo.is.connected(mongo) == TRUE) {
  mongo.insert(mongo, "taxi.yellowsamples", list(yellow_bson))
}

```

## Sample data was stored in MongoDb using mongoimport

### Connect to MongoDB and get some basic statistics

```{r Connect to MongoDB and get some basic statistics}

mongoyel <- mongo(collection = "yellowsample", db = "taxi", url = "mongodb://localhost", verbose = TRUE)
mongoyel$count()
mongoyel$find('{"VendorID": 1}', limit = 1)

mongogr <- mongo(collection = "greensample", db = "taxi", url = "mongodb://localhost", verbose = TRUE)
mongogr$count()
mongogr$find('{"VendorID": 1}', limit = 1)
```

Number of rows in yellowsample
1315826
 
Number of rows in greensample
178685

These amount of rows proved to be a challenge, hence the sample was sampled again to get the rows under 300,000. 15% of this sample was taken for an overall sample of 1.5% of the data for the month of May.

### Sample the sample - this time using the mongolite package

```{r}
# Load sample data from MongoDB
yelcon <- mongo(collection = "yellowsample", db = "taxi", url = "mongodb://localhost", verbose = TRUE)
yellow <- yelcon$find('{}')

grncon <- mongo(collection = "greensample", db = "taxi", url = "mongodb://localhost", verbose = TRUE)
green <- grncon$find('{}')

# Sample the sample
set.seed(201505)
yelsamp <- yellow[sample(1:dim(yellow)[1], 197374), ]

set.seed(052015)
grnsamp <- green[sample(1:dim(green)[1], 26803), ]

yel2 <- yelsamp
yelsamp$RateCodeID <- factor(yelsamp$RateCodeID, levels = c(1, 2,3 ,4 ,5 ,6, 99), labels = c("Standard rate", "JFK", "Newark", "Nassau or Westchester", "Negotiated fare", "Group ride", "Unknown"))

grnsamp$RateCodeID <- factor(grnsamp$RateCodeID, levels = c(1, 2,3 ,4 ,5 ,6, 99), labels = c("Standard rate", "JFK", "Newark", "Nassau or Westchester", "Negotiated fare", "Group ride", "Unknown"))

grnsamp$Ehail_fee <- NULL
colnames(yelsamp) <- tolower(colnames(yelsamp))
colnames(grnsamp) <- tolower(colnames(grnsamp))

yel3 <- yelsamp[c("ratecodeid", "trip_distance")]
grn3 <- grnsamp[c("ratecodeid", "trip_distance")]


totalsamp <- bind_rows(yel3, grn3)




taxidata$ratecodeid = totalsamp$ratecodeid

```


*** MIGHT NOT NEED THIS
### Backup final samples to MongoDB
Most of the editing of these files will be done in MOngoDB. This is because the file the yellow taxi data for May is over 2GB. Therefore goinf forward it is easier to manipulate these files in the database given the computing resiurces available.

```{r Backup sample files, echo=FALSE}

# Setup new connection to MongoDb
yelsamp_write <- mongo(collection = "yelsamp", db = "taxi", url = "mongodb://localhost", verbose = TRUE)
stream_out(yelsamp, file("yelsamp.ndjson"), pagesize = 1000)
yelsamp_write$import(file("yelsamp.ndjson"), bson = FALSE)


# Test that the JSON file was imported to MongoDB
yelsamp_write$count()
yelsamp_write$find('{"VendorID": 1}', limit = 1)

#yelsamp <- yelsamp_write$find('{}')

grnsamp_write <- mongo(collection = "grnsamp", db = "taxi", url = "mongodb://localhost", verbose = TRUE)
stream_out(grnsamp, file("grnsamp.ndjson"), pagesize = 1000)
grnsamp_write$import(file("grnsamp.ndjson"), bson = FALSE)

# Test that the JSON file was imported to MongoDB
grnsamp_write$count()
grnsamp_write$find('{"VendorID": 1}', limit = 1)

#grnsamp <- grnsamp_write$find('{}')

```


### Drop empty columns from green taxis

```{r Drop empty columns, echo= FALSE}

grnsamp$Ehail_fee <- NULL

```


### Change green column names to lowercase, drop unused column and rename and simplify field names 
We want both samples to have the same field names

```{r Change column names to lower case, echo=FALSE}

colnames(yelsamp) <- tolower(colnames(yelsamp))
colnames(grnsamp) <- tolower(colnames(grnsamp))

# Drop E-hail column
grnsamp$ehail_fee <- NULL

# Rename and simplyfy column names

```

### Change date information using dplyr and lubridate

```{r Change date information, echo+FALSE}

changetodate <- function(chardate, pref) {
  dimdate = length(chardate)
  
  #Create empty data frame
  getdate <- data.frame(numeric(0), numeric(0), numeric(0), numeric(0), numeric(0), numeric(0), numeric(0), numeric(0), numeric(0))
  
  # Initialize data frame
  getdate[dimdate, 9] <- 0
  
  # Name data frame columns
  colnames(getdate) <- c(paste0(pref, "datetime"), paste0(pref,"year"), paste0(pref,"month"), paste0(pref,"day"), paste0(pref,"hour"), paste0(pref,"minute"), paste0(pref,"second"), paste0(pref,"yday"), paste0(pref, "wday"))
  
  # Extract time series info using lubridate package
  getdate[,1] <- ymd_hms(chardate, tz = "America/New_York")
  getdate[,2] <- year(getdate[,1])
  getdate[,3] <- month(getdate[,1])
  getdate[,4] <- day(getdate[,1])
  getdate[,5] <- hour(getdate[,1])
  getdate[,6] <- minute(getdate[,1])
  getdate[,7] <- second(getdate[,1])
  getdate[,8] <- yday(getdate[,1])
  getdate[,9] <- wday(getdate[,1], label = TRUE)
  
  getdate
  
}

gnew_pckdates <- changetodate(grnsamp$lpep_pickup_datetime, "pck")
gnew_dropdates <- changetodate(grnsamp$lpep_dropoff_datetime, "drp")

gnew <- bind_cols(gnew_pckdates, gnew_dropdates)

ynew_pckdates <- changetodate(yelsamp$tpep_pickup_datetime, "pck")
ynew_dropdates <- changetodate(yelsamp$tpep_dropoff_datetime, "drp")

ynew <- bind_cols(ynew_pckdates, ynew_dropdates)

```

### ### Rename columns of yellow and green taxi data. Create new column for yellow taxis.

```{r Rename columns, echo=FALSE}

# Rename columns in both data frames to match
yelsamp <-  yelsamp %>%
                    dplyr::rename(pickup_datetime = tpep_pickup_datetime, dropoff_datetime = tpep_dropoff_datetime)

grnsamp <-  grnsamp %>%
                    dplyr::rename(pickup_datetime = lpep_pickup_datetime, dropoff_datetime = lpep_dropoff_datetime, trip_type = trip_data)


# Create column in yellow taxi data frame called trip_type and populate with 3
yelsamp$trip_type <- 3

# Create column in yellow and green taxi data frames called type and populate with yellow and green respectively for the type of taxi.

yelsamp$type <- "yellow"
grnsamp$type <- "green"

```

### Combine new times series colums and yellow and greeen taxi rows to form one data frame

```{r Combine all new data. echo=FALSE}

# Add new columns to yellow and green taxi data frame
yeldata <- bind_cols(yelsamp, ynew)
grndata <- bind_cols(grnsamp, gnew)

# Combine yellow and green taxi data to form one data frame
taxidata <- bind_rows(yeldata, grndata)

# Create id column in taxidata data.frame
dim(taxidata)
id <- 1:dim(taxidata)[1]

taxidata$id <- id

```

### Write taxi data into MongoDB and CSV

```{r Write data to MongoDB and CSV, echo=FALSE}

# Convert taxi data into JSON format

#writeLines (taxiJSON, "taxiday.ndjson")

taxi <- mongo(collection = "taxinyc", db = "taxi", url = "mongodb://localhost", verbose = TRUE)
stream_out(taxidata, file("taxi3_loc.ndjson"), pagesize = 1000)
taxi$import(file("taxi.ndjson"), bson = FALSE)

taxi$count()

```

### Create factor columns and add new trip duration column

```{r Factor columns, echo=FALSE}

# Convert character fields to Date

cNames <- c("pckdatetime", "drpdatetime")

taxicols <- as.data.frame(lapply(taxidata[cNames], strptime, format = "%Y-%m-%d %H:%M:%S"))
taxidata[cNames] <- taxicols

# Convert numeric categories to characters and then Factors
factorcols <- c("vendorid", "ratecodeid", "store_and_fwd_flag", "payment_type", "trip_type", "type")
taxidata.f <- taxidata %>%
              dplyr::select(one_of(factorcols))

taxidata.f$vendorid <- factor(taxidata.f$vendorid, levels = c(1, 2), labels = c("Creative Mobile Tech", "Verifone"))

taxidata.f$ratecodeid <- factor(taxidata.f$ratecodeid, levels = c(1, 2,3 ,4 ,5 ,6, 99), labels = c("Standard rate", "JFK", "Newark", "Nassau or Westchester", "Negotiated fare", "Group ride", "Unknown"))

taxidata.f$store_and_fwd_flag <- factor(taxidata.f$store_and_fwd_flag, levels = c("Y", "N"))

taxidata.f$payment_type <- factor(taxidata.f$payment_type, levels = c(1, 2, 3, 4, 5, 6), labels = c("Credit card", "Cash", "No charge", "Dispute", "Unknown", "Voided"))

taxidata.f$trip_type <- factor(taxidata.f$trip_type, levels = c(1, 2, 3), labels = c("Street-hail", "Dispatch", "No Info"))

taxidata.f$type <- factor(taxidata.f$type, levels = c("yellow", "green"))

taxidata[factorcols] <- taxidata.f

# Create an easy to use trip_duration field. Measure in miuntes. Add as new column to taxi data
trip_duration <- time_length(interval(taxidata$pckdatetime, taxidata$drpdatetime), "minute")
taxidata$trip_duration <- trip_duration

```

### Save to CSV File

```{r Save to CSV, echo=FALSE}

write.csv(taxidata, "taxidata5.csv")
#write.csv(taxidata, "taxidata5_loctimedestwkday.csv")

```

